SAT Solving for Vesicle Traffic Systems in Cells

### 2017-07-13 paper/bio.tex
changes

%Again about two pages long

%\item \mukund{Detailed BIO}
\textbf{The cell as a transport graph:} We consider a cell to be a collection of
compartments (nodes) and vesicles (edges), thus defining a transport graph.
Compartments represent the large membrane-bound organelles, which are
connected by constant fluxes of small transport vesicles. Every compartment
or vesicle has a set of molecules associated with it. These include SNAREs
and other biomolecules that regulate membrane traffic, or that play the
role of passive cargo. Here we consider an abstract set of molecules and do
not explicitly associate them with particular protein varieties.

\textbf{Molecular flows and steady state:} The total amount of each molecular type
on each compartment can increase or decrease: this is due to gain or loss
of that molecular type driven by incoming or outgoing vesicles. Each edge
is thus associated with a flux of all the molecular types carried by the
corresponding vesicle. We assume the cell is in a steady state where each
compartments composition does not vary over time, meaning that all
incoming and outgoing fluxes are balanced for each molecular type at each
compartment.

\textbf{Vesicle targeting driven by molecular interactions:} The model specification
includes a description of how molecular properties influence vesicle
transport. A vesicle can only contain a set of molecules that are present
on its source compartment. Once a vesicle has budded out of the source, the
molecules it carries determine its properties. In particular, for any given
pair of a vesicle and a compartment, the set of molecules that label the
former and latter determine whether the vesicle will fuse to that
compartment. Biophysically, fusion requires a direct physical interaction
between at least one molecular type on the vesicle and one molecular type
on the compartment. The list of molecular pairs that can drive a fusion
event is given an a fusion pairing matrix.

\textbf{Molecular regulation:} The final layer of our framework involves how the
molecules are regulated. We assume that for fusion to occur, the pair of
molecular types involved on the vesicle and compartment must both be in an
“active” state. Whether these molecules are active or inactive depends on
the remaining molecules found on the vesicle or compartment, respectively.
This embodies the biological fact that other molecules can regulate the
fusion-driving molecules. We test many different versions of molecular
regulation. Most generally, the activity state of a given molecule can be a
Boolean function of all the molecular types on a compartment or vesicles.
We have also tested a particularly simple regulation mechanism in which two
molecules that can pair to drive fusion inhibit one another. This is
motivated by the idea that pairing must generate an inactive bi-molecular
complex.

\textbf{Synthesis:} We now combine all the above ingredients. Given a particular
transport graph, a particular labeling of all the compartments and edges,
a particular fusion pairing matrix, and a particular regulatory model we do
the following.
\begin{enumerate}
\item We determine which molecules are ``active" on every
compartment or vesicle.
\item or every vesicle fusing to a compartment, we
determine whether there exists an active pair (one molecule on the vesicle,
one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the
compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We verify that every molecular type entering a compartment also leaves the compartment,
and also that every molecular type entering a set of compartments also
leaves that set; this is the steady state condition. This tells us that the
particular graph and molecular labeling does represent an allowed steady
state configuration of a VTS.
\end{enumerate}

\begin{itemize}
\item \srivas{Detailed BIO to Graph problem}
\end{itemize}

%he whole cell should be in stable condition i.e flow of the molecule is balanced at every point and globally. Each molecule leaving its source node on a vesicle should come back in a cycle.

The following terms must be defined in this section
\begin{itemize}
\item Vesicular traffic network.
\item Recycling of molecules, steady state.
\item Maximum connectivity [LGC, Least guarantee connectivity] and minimal connective [LRC (least required connectivity)] (rather than necc and suff cond) required by the graph. 
\item Connected-ness. 3 and 4.
\item Pairing mechanism (matrix).
\item Fusion rules, inhibition/regulation mechanism, Pairing inhibition.
\item Potential to connect.
\item Property of interest.
\item Reachability definition.
\end{itemize}

>>>>

%Again about two pages long
%
In this section, we will describe the model of VTSs and the properties
relevant to the biologists.

\textbf{The cell as a transport graph:} We consider a cell to be a
collection of compartments (nodes) and vesicles (edges), thus defining
a transport graph.
%
Compartments represent the large membrane-bound organelles, which are
connected by constant fluxes of small transport vesicles.
%
Every compartment or vesicle has a set of molecules associated with
it.
%
These include SNAREs and other biomolecules that regulate membrane
traffic, or that play the role of passive cargo.
%
Here we consider an abstract set of molecules and do
not explicitly associate them with particular protein varieties.

\textbf{Molecular flows and steady state:} The total amount of each
molecular type on each compartment can increase or decrease: this is
due to gain or loss of that molecular type driven by incoming or
outgoing vesicles.
%
Each edge is thus associated with a flux of all the molecular types
carried by the corresponding vesicle.
%
We assume the cell is in a steady state where each compartments
composition does not vary over time, meaning that all incoming and
outgoing fluxes are balanced for each molecular type at each
compartment.

\textbf{Vesicle targeting driven by molecular interactions:} The model
specification includes a description of how molecular properties
influence vesicle transport.
%
A vesicle can only contain a set of molecules that are present
on its source compartment.
%
Once a vesicle has budded out of the source, the molecules it carries
determine its properties.
%
In particular, for any given pair of a vesicle and a compartment, the
set of molecules that label the former and latter determine whether
the vesicle will fuse to that compartment.
%
Biophysically, fusion requires a direct physical interaction between
at least one molecular type on the vesicle and one molecular type on
the compartment.
%
The list of molecular pairs that can drive a fusion event is given an
a fusion pairing matrix.

\textbf{Molecular regulation:} The final layer of our framework
involves how the molecules are regulated.
%
We assume that for fusion to occur, the pair of molecular types
involved on the vesicle and compartment must both be in an
“active” state.
%
Whether these molecules are active or inactive depends on the
remaining molecules found on the vesicle or compartment, respectively.
%
This embodies the biological fact that other molecules can regulate the
fusion-driving molecules.
%
VTSs also require that if molecule is involved in fusion of some
vesicle and compartment, it must not be possible that the molecule 
can potentially fuse with some molecule in some other compartment.
%

We test many different versions of molecular
regulation.
%
Most generally, the activity state of a given molecule can be a
Boolean function of all the molecular types on a compartment or vesicles.
We have also tested a particularly simple regulation mechanism in which two
molecules that can pair to drive fusion inhibit one another.
%
This is
motivated by the idea that pairing must generate an inactive bi-molecular
complex.

\textbf{Synthesis:} We now combine all the above ingredients.
%
Given a particular
transport graph, a particular labeling of all the compartments and edges,
a particular fusion pairing matrix, and a particular regulatory model we do
the following.
\begin{enumerate}
\item We determine which molecules are ``active" on every
compartment or vesicle.
\item or every vesicle fusing to a compartment, we
determine whether there exists an active pair (one molecule on the vesicle,
one on the compartment) which drives that fusion event.
\item For every vesicle-compartment pair where the vesicle does not fuse to the
compartment, we verify that there is no pairing of active molecules on the
vesicle and compartment that could drive their fusion.
\item We verify that every molecular type entering a compartment also leaves the compartment,
and also that every molecular type entering a set of compartments also
leaves that set; this is the steady state condition.
%
This tells us that the
particular graph and molecular labeling does represent an allowed steady
state configuration of a VTS.
\end{enumerate}

Given the above conditions, the biologists search for VTSs that are
not $k$-connected, i.e., every pair of compartments remain reachable
after dropping $k$ vesicles.\ashu{why asking this question}
\ankit{Maximum connectivity [LGC, Least guarantee connectivity] and minimal connective [LRC (least required connectivity)] (rather than necc and suff cond) required by the graph.}

### 2017-07-13 paper/encoding.tex
table-update
added-pairing matrix
added dia
changes

We will also consider several variations of the model.
%
For example, unique edge between two nodes, activity of molecules is
not constrained by $\nodef$ and $\edgef$, and \ashu{other
variations?}.
%

@@

\paragraph{Search problem}

Based upon earlier discussion, we need to answer the following search
question for ......\ashu{match with the earlier text}
For a given $k$, size $\nu$, and molecule number $\mu$,
we are searching for well-structured, stable, and well-fused VTS
$G = (\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$ such that
$|\nodes| = \nu$, $|\mols = \mu|$, and
$G$ is not $k$-connected.

@@

Condition on $p_{kk'}$. \ashu{why???} Diagonal blocks should be all 0's.

\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V7}\label{eq:c3}
\end{align}

@@

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\item Version F with additional constraint that every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
\item Version F but self-edges are allowed.
\end{enumerate}

@@

>>>>

We will also consider several variations of the model.
%
For example, unique edge between two nodes, activity of molecules is
not constrained by $\nodef$ and $\edgef$, etc.
%

@@

\paragraph{Search problem}
Based upon earlier discussion, we need to answer the following search
question among VTSs.
For a given $k$, size $\nu$, and molecule number $\mu$,
we are searching for well-structured, stable, and well-fused VTS
$G = (\nodes,\mols,\edges,\nlabel,\pairs,\edgef,\nodef)$ such that
$|\nodes| = \nu$, $|\mols = \mu|$, and
$G$ is not $k$-connected.

@@

Since a molecule cannot be both SNARE and tether,
we set diagonal blocks in pairing matrix to be 0's.
It reduces the search space.
% Condition on $p_{kk'}$. \ashu{why???} .
\begin{align}
  \bigwedge\limits_{(x < M/2 \, \land  \, y < M/2) \lor  (x >= M/2 \land y >= M/2)} \neg p(x,y)
  \tag{V7}\label{eq:c3}
\end{align}

@@

\begin{enumerate}[label=\Alph*]
\item Every present molecule is considered to be active.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules. 
\item Activity of molecules on the edges is based on Boolean function of presence of other molecules
\item Activity of molecules both on the edges and nodes is based on Boolean function of presence of other molecules.
\item Activity of molecules on the edges is driven by pairing inhibition.
\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition.
\item Version F with additional constraint that every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other molecules and on edge by pairing inhibition with every edge is distinct.
%\item Activity of molecules on the nodes is based on Boolean function of presence of other moleculess and on edge by pairing inhibition with self-edges are allowed.
%\item Version F but self-edges are allowed.
\end{enumerate}

@@

\input{pm-1-matrix}
\input{v2-dia}

### 2017-07-13 paper/intro.tex
changes

The conservation of basic molecular processes across all living cells, including the genetic code, suggests that life on Earth arose only once: all extant organisms are presumed to trace their lineages back to a last universal common ancestor (LUCA). However, there are still vast differences in cellular and molecular architecture among the three kingdoms of life: bacteria, archaea, and eukaryotes \cite{embley2006eukaryotic}. Eukaryotic cells have larger genomes and have more complex cellular architectures than the prokaryotic bacteria and archaea. In particular, the cytoplasm of all eukaryotic cells is broken up into distinct membrane-bound locations known as organelles: the nuclear membrane, endoplasmic reticulum (ER), Golgi apparatus, vacuoles and lysosomes are examples of these. Cargo is moved between these organelles in small membrane-bound transporters known as vesicles \cite{stenmark2009rab}.

This organization resembles a transport logistics network, and is commonly known as the membrane traffic system. Membrane traffic underpins nearly every aspect of eukaryotic cellular physiology, including human cellular physiology. Defects in membrane traffic lead to a variety of disorders or even cell death \cite{stenmark2009rab}. Understanding how membrane traffic functions is therefore one of the central challenges of cell biology. Some progress has been made, as cell biologists have assembled a “parts list” of molecules that drive vesicle traffic \cite{dacks2007evolution}. The essential processes involve the creation of vesicles loaded with cargo from source organelles, known as “budding”, and the depositing of those vesicles and cargo into target organelles, known as “fusion” \cite{munro2004organelle}. Budding is regulated by proteins known as coats and adaptors that select cargo. Fusion is regulated by proteins known as SNAREs and tethers, that ensure that vesicles fuse to the correct target \cite{mani2016stacking}. Although we know a great deal about molecular-level details, there has been very little work done on how such processes are integrated across scales to build the entire vesicle traffic system. We have recently attempted to use ideas of graph theory to address this question \cite{mani2016stacking,shukla}.

The SNARE proteins and the constraints that SNARE-to-SNARE regulations impose on feasibile behaviors of vesicular systems are the focus of our current work. Abstractly, we can think of these proteins as labels that are collected from the source compartment and taken to the target compartment on vesicles. There is a corresponding set of label molecules on the target compartments. If the two sets of labels (on vesicles and targets) are compatible, then the vesicle will fuse. The question is: does this physical picture place any constrains on the global topology of the traffic graph? We have previously shown that one informative constraint is graph connectivity \cite{shukla}.  The intricacies of SNARE regulation are largely unknown. How stringent SNARE regulation is can determine whether graph topologies of particular level of connectivity are biologically realizable.  For example, if SNAREs are completely unregulated and thus always active, they cannot be used to generate a traffic network. If these SNAREs are regulated by relatively simple rules then only highly connected traffic networks can be generated using them.   The question is can we work toward a predictive analytical framework of vesicle traffic?

The analysis of vesicle traffic systems is a difficult problem because of the combinatorial scaling of possible traffic topologies and regulatory rules.  Previous analyses of vesicle traffic networks dealt with this combinatorial explosion by using a sampling approach: simulating randomly generated regulatory rules and taking averages over many samples \cite{mani2016stacking}.  The outcome of such simulations is necessarily statistical rather than rigorous and precise. To make definite predictions about vesicle traffic network topologies, we need to efficiently check useful system properties over all possible graphs and all possible variations of rules, preferably without enumerating them.  Formal verification tools, such as model checkers \cite{clarke1996symbolic, biere2003bounded, clarke2008birth, cimatti2000nusmv, holzmann1997model} and SAT solvers \cite{moskewicz2001chaff,een2004extensible} serve precisely this purpose. These tools encode the problem symbolically, and check various properties without explicit enumeration of the graphs up to large but finite size of the models.
%These tools are central to formal verification methods in computer science, and are increasingly being applied to understand biological systems, including gene regulatory networks [16-18].

In a recent work, we employed the model checker CBMC \cite{CKY03, ckl2004} meant to analyze ANSI C programs for studying this problem.  In this work, vesicular systems and SNARE regulations were modeled as C code mostly using arrays to represent graphs and boolean tables.  Queries about graph connectivity requirements for different variations of SNARE Regulations were modeled as logical assertions to be checked by the model checker.  CBMC uses its own built-in encodings to reduce the analysis problem to boolean satisfiability problem which is solved using SAT solvers.  This work was successful in demonstrating the feasibility of this technology to this domain.  However, the scalability of this approach was limited (up to vesicular systems with 7-8 nodes) due to the following reasons: 
\begin{enumerate}
\item The encoding used to model vesicular systems was not fine-tuned to the structure of the problem.
\item The queries to be solved involve quantified formulas, i.e., we needed to solve queries of the form ``\emph{for all} graphs does  there \emph{exist} SNARE regulation constraints that satisfy biological steady state.
\end{enumerate}
As CBMC does not directly support solving such quantified formulas, we had to use a technique that combined non-determinism and partial enumeration that is not as efficient.\newline
%\item \srivas{BIO to Graph problem}

%\item \srivas{Graph to previous encoding}

%\end{itemize}

% \noindent

% What is vesicle traffic systems

% Graph properties of the systems

% Search queries; why bio imp

% Using SAT/QBF solvers to encode

%
\textbf{Vesicle traffic systems(VTS)} In this paper, we have developed a novel SAT encoding for
searching a VTS that satisfy the given properties.
%
We improved the encoding of some of the conditions for VTSs.
%
In the earlier implementation, reachability was encoded by
enumeration of paths.
%
In our encoding, we have recursively defined the reachability condition
such that that we avoid the exponential enumeration.
%
We have also encoded $k$-connectivity in the similar fashion and
avoided a blowup in the formula size.
%

>>>>

The conservation of basic molecular processes across all living cells,
including the genetic code, suggests that life on Earth arose only
once: all extant organisms are presumed to trace their lineages back
to a last universal common ancestor (LUCA).
%
However, there are still vast differences in cellular and molecular
architecture among the three kingdoms of life: bacteria, archaea, and
eukaryotes~\cite{embley2006eukaryotic}.
%
Eukaryotic cells have larger genomes and have more complex cellular
architectures than the prokaryotic bacteria and archaea.
%
In particular, the cytoplasm of all eukaryotic cells is broken up into
distinct membrane-bound locations known as organelles: the nuclear
membrane, endoplasmic reticulum (ER), Golgi apparatus, vacuoles and
lysosomes are examples of these.
%
Cargo is moved between these organelles in small membrane-bound
transporters known as vesicles \cite{stenmark2009rab}.

This organization resembles a transport logistics network, and is
commonly known as the membrane traffic system.
%
Membrane traffic underpins nearly every aspect of eukaryotic cellular
physiology, including human cellular physiology.
%
Defects in membrane traffic lead to a variety of disorders or even
cell death \cite{stenmark2009rab}.
%
Understanding how membrane traffic functions is therefore one of the
central challenges of cell biology.
%
Some progress has been made, as cell biologists have assembled a
“parts list” of molecules that drive vesicle
traffic~\cite{dacks2007evolution}.
%
The essential processes involve the creation of vesicles loaded with
cargo from source organelles, known as “budding”, and the depositing
of those vesicles and cargo into target organelles, known as “fusion”
\cite{munro2004organelle}.
%
Budding is regulated by proteins known as coats and adaptors that
select cargo.
%
Fusion is regulated by proteins known as SNAREs and tethers, that
ensure that vesicles fuse to the correct target
\cite{mani2016stacking}.
%
Although we know a great deal about molecular-level details, there has
been very little work done on how such processes are integrated across
scales to build the entire vesicle traffic system (VTS).
%
We have recently attempted to use ideas of graph theory to address
this question \cite{mani2016stacking,shukla}.

The SNARE proteins and the constraints that SNARE-to-SNARE regulations
impose on feasibile behaviors of VTSs are the focus of
our current work.
%
Abstractly, we can think of these proteins as labels that are
collected from the source compartment and taken to the target
compartment on vesicles.
%
There is a corresponding set of label molecules on the target compartments.
%
If the two sets of labels (on vesicles and targets) are compatible,
then the vesicle will fuse.
%
The question is: does this physical picture place any constrains on
the global topology of the traffic graph? We have previously shown
that one informative constraint is graph connectivity \cite{shukla}.
%
The intricacies of SNARE regulation are largely unknown.
%
How stringent SNARE regulation is can determine whether graph
topologies of particular level of connectivity are biologically
realizable.
%
For example, if SNAREs are completely unregulated and thus always
active, they cannot be used to generate a traffic network.
%
If these SNAREs are regulated by relatively simple rules then only
highly connected traffic networks can be generated using them.
%
The question is can we work toward a predictive analytical framework
of VTSs?

The analysis of VTSs is a difficult problem because
of the combinatorial scaling of possible traffic topologies and
regulatory rules.
%
Previous analyses of VTSs dealt with this combinatorial explosion by
using a sampling approach: simulating randomly generated regulatory
rules and taking averages over many samples~\cite{mani2016stacking}.
%
The outcome of such simulations is necessarily statistical rather than
rigorous and precise.
%
To make definite predictions about VTS topologies, we need to
efficiently check useful system properties over all possible graphs
and all possible variations of rules, preferably without enumerating
them.
%
Formal verification tools, such as model checkers
\cite{clarke1996symbolic, biere2003bounded, clarke2008birth,
cimatti2000nusmv, holzmann1997model} and SAT solvers
\cite{moskewicz2001chaff,een2004extensible} serve precisely this
purpose.
%
These tools encode the problem symbolically, and check various
properties without explicit enumeration of the graphs up to large but
finite size of the models.

%These tools are central to formal verification methods in computer science, and are increasingly being applied to understand biological systems, including gene regulatory networks [16-18].

In a recent work, we employed the model checker CBMC \cite{CKY03,
ckl2004} meant to analyze ANSI C programs for studying this problem.
%
In this work, VTSs and SNARE regulations were modeled as C code mostly
using arrays to represent graphs and boolean tables.
%
Queries about graph connectivity requirements for different variations
of SNARE Regulations were modeled as logical assertions to be checked
by the model checker.
%
CBMC uses its own built-in encodings to reduce the analysis problem to
boolean satisfiability problem which is solved using SAT solvers.
%
This work was successful in demonstrating the feasibility of this
technology to this domain.
%
However, the scalability of this approach was limited (up to VTSs
with 7-8 nodes) due to the following reasons:
\begin{enumerate}
\item The encoding used to model VTSs was not fine-tuned
to the structure of the problem.
\item The queries to be solved involve quantified formulas, i.e., we
needed to solve queries of the form ``\emph{for all} graphs does there
\emph{exist} SNARE regulation constraints that satisfy biological
steady state.
\end{enumerate} As CBMC does not directly support solving such
quantified formulas, we had to use a technique that combined
non-determinism and partial enumeration that is not as
efficient.\newline

%
In this paper, we have developed a novel SAT encoding for
searching a VTS that satisfy the given properties.
%
We improved the encoding of some of the conditions for VTSs.
%
In the earlier implementation, reachability was encoded by
enumeration of paths.
%
In our encoding, we have recursively defined the reachability condition
such that that we avoid the exponential enumeration.
%
We have also encoded $k$-connectivity in the similar fashion and
avoided a blowup in the formula size.
%

### 2017-07-13 paper/pm-1-matrix.tex
added dia
changes

\begin{figure}
\centering
\begin{tikzpicture}
\matrix [matrix of math nodes,left delimiter=(,right delimiter=),row sep=0.5cm,column sep=0.5cm] (m) {
\times & \times & \times &\times&\times&\times & 1 & \times\\
\times & \times & \times&\times&\times&1& \times &\times \\
\times & \times & \times&\times&1&\times&\times &\times\\
\times & \times & \times&\times & \times & \times &\times & \times\\
 \times & \times & \times & \times & \times & \times & \times & \times\\
 \times & \times & \times &\times & \times & \times & \times & \times\\
 \times & \times & \times & \times & \times & \times & \times & \times\\
 \times & \times & \times & \times & \times & \times&\times & \times\\};

@@

\node[above=10pt of m-1-1] (top-1) {$M_1$};
\node[above=10pt of m-1-2] (top-2) {$M_2$};
\node[above=10pt of m-1-3] (top-3) {$M_3$};
\node[above=10pt of m-1-4] (top-4) {$M_4$};
\node[above=10pt of m-1-5] (top-5) {$M_5$};
\node[above=10pt of m-1-6] (top-6) {$M_6$};
\node[above=10pt of m-1-7] (top-7) {$M_7$};
\node[above=10pt of m-1-8] (top-8) {$M_8$};

@@

\node[rectangle,left delimiter=\{] (del-left-1) at ($0.5*(left-1.east) +0.5*(left-4.east)$) {\tikz{\path (left-1.north east) rectangle (left-4.south west);}};
\node[left=10pt] at (del-left-1.west) {$Q-Snares$};
\node[rectangle,left delimiter=\{] (del-left-2) at ($0.5*(left-5.east) +0.5*(left-8.east)$) {\tikz{\path (left-5.north east) rectangle (left-8.south west);}};
\node[left=10pt] at (del-left-2.west) {$R-Snares$};

@@

%\foreach \element in {m-1-1,m-6-5,m-5-3}{
%\highlight[pink]{\element}{\element}
%}
\foreach \element in {m-1-7,m-2-6,m-3-5}{
\highlight[blue!50]{\element}{\element}
}
\end{pgfonlayer}

\end{tikzpicture}

@@

>>>>

\begin{figure}
\centering
\begin{minipage}{0.45\linewidth}
  \hspace{-4ex}
\begin{tikzpicture}
\matrix [matrix of math nodes,left delimiter=(,right delimiter=),row sep=0.16cm,column sep=0.1cm] (m) {
\times & \times & \times &\times&\times&\times & \times & \times\\
\times & \times & \times&\times&\times&\times& 1 &\times \\
\times & \times & \times&\times&\times&1&\times &\times\\
\times & \times & \times&\times & 1 & \times &\times & \times\\
 \times & \times & \times & \times & \times & \times & \times & \times\\
 \times & \times & \times &\times & \times & \times & \times & \times\\
 \times & \times & \times & \times & \times & \times & \times & \times\\
 \times & \times & \times & \times & \times & \times&\times & \times\\};

@@

\node[above=4pt of m-1-1] (top-1) {$M_1$};
\node[above=4pt of m-1-2] (top-2) {$M_2$};
\node[above=4pt of m-1-3] (top-3) {$M_3$};
\node[above=4pt of m-1-4] (top-4) {$M_4$};
\node[above=4pt of m-1-5] (top-5) {$M_5$};
\node[above=4pt of m-1-6] (top-6) {$M_6$};
\node[above=4pt of m-1-7] (top-7) {$M_7$};
\node[above=4pt of m-1-8] (top-8) {$M_8$};

@@

\node[rectangle,left delimiter=\{] (del-left-1) at ($0.5*(left-1.east) +0.5*(left-4.east)$) {\tikz{\path (left-1.north east) rectangle (left-4.south west);}};
\node[left=15pt,rotate=90,xshift=9mm] at (del-left-1.west) {$Q-Snares$};
\node[rectangle,left delimiter=\{] (del-left-2) at ($0.5*(left-5.east) +0.5*(left-8.east)$) {\tikz{\path (left-5.north east) rectangle (left-8.south west);}};
\node[left=15pt,rotate=90,xshift=9mm] at (del-left-2.west) {$R-Snares$};

@@

\foreach \element in {m-1-7,m-3-8,m-5-1,m-5-2,m-5-3,m-5-4,m-6-1,m-6-2,m-6-3,m-6-4,m-7-1,m-7-2,m-7-3,m-7-4,m-8-1,m-8-2,m-8-3,m-8-4}{
\highlight[pink]{\element}{\element}
}
\foreach \element in {m-2-7,m-3-6,m-4-5}{
\fhighlight[blue!10]{\element}{\element}
}
\end{pgfonlayer}

\end{tikzpicture}  
\end{minipage}
\begin{minipage}{0.45\linewidth}
  \tikzset{
    ln/.style  = { draw, thick, fill=black, circle, inner sep=0.3mm},
    sqloc/.style  = { draw, thick, inner sep=1.0mm },
    loc/.style    = { draw, circle, thick, inner sep=1.0mm },
  }
\begin{tikzpicture}
  \node[draw] (n0) {$n_1:\{0,1,2,3,4\}$};
  \node[draw,below left of=n0] (n1) {$n_1:\{0,1,2,3,4\}$};
  \node[draw] (n2) {$n_1:\{0,1,2,3,4\}$};

    \entity[yshift=-20mm] (E) {$n_1$ $[m_0 m_1 m_2 m_3 m_4]$};
    % \entity (E) {Node1\\[]};
    \binary[left=of E.north west, unique=2] (h) {};

       % \binary[left=of E.north west, unique=1-2] (w) {};

    \binary[left=of E.south west, unique=1-2](i) [below=6mm of h]{};
    \node[role name, at=(i.north),anchor=south] {[$m_{0} m_{1} m_{3}$]};
    \plays[] (E) to (i.east);
   % \plays(h) to (i);
%         label=below:\ormleft{uses] (u) {};
\node[role name,
at=(h.north),anchor=south] {[$m_{0} m_{1}$]};

    \entity[left=of h] (Rank) {$n_0:m_0 m_1 m_2 m_3 m_6$};

    \plays[mandatory] (E) to (h.east);
    \plays (h) to (Rank) (E);
    \plays[mandatory] (Rank) to (i.west);
    %to (u.east) (u) to (Car);

    \binary[right=of E.north east] (b) {};
    \node[role name,
at=(b.north),anchor=south] {[$m_{0} m_{1}$]};
       % \binary[right=of E.north east] (dd) {};

    \binary[right=of E.south east, unique,] (i) {};
    \node[role name,
at=(i.north),anchor=south] {[$m_{1} m_{2} m_{3}$]};
\binary[unique=1-2,skip unique=1-2] (xyz) [above=6mm of h] {};
% describes blue label above h Left uppermost
\node[role name,
at=(xyz.north),anchor=south] {[$m_{0} m_{1} m_{3}$]};

      \plays (Rank) to (xyz.west); 
    \plays[mandatory] (E) to (xyz.east);

    \entity[right=1.8 of E] (Date) {$n_2: m_1 m_2 m_3 m_6 m_8$};

    \plays[mandatory] (E) to (b.west); 
    \plays (E) to (i.west);
    \plays (b.east) to (Date);
    \plays[mandatory] (Date) to (i.east);

%    \entity[label={[constraint]\#=$n$}] {E};
%\draw (Rank) to node[vroles,label=has,label=below:\ormup{of}]{} (E);
%\node[constraint,anchor=north west,inner ysep=0]
%at (h.) {\ormbraces{a, b, c}};
%\draw (E) to node[roles,
%label=below:drives]{} (Rank);

\node[roles=2][,unique=1-2,skip unique=2] (abc) [below=11mm of E] {};
\node[role name,
at=(abc.north),anchor=south] {[$m_{2} m_{3}$]};
\plays[mandatory] (Rank.south) -- (abc.west);  
\plays (abc.east) -- (Date.south);

  %  \node[role name, anchor=south west] at (b.east) {[birthdate]};
  %  \node[role name, anchor=north west] at (i.east) {[hiredate]};

%     \rules at (-.4, -2) {
%          \node[rule=1] {{$m_{2}$} $m_{3}$};\\
%      };
\end{tikzpicture}
\end{minipage}

@@

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
